---
layout: post
title: "You wanna PEESE of d's?"
date: April 18, 2017
tags: [meta-analysis]
---

Publication bias---and more generally, outcome reporting bias---is now recognized as a critical threat to the validity of findings from research syntheses. In the areas with which I am most familiar (education and psychology), it has become more or less a requirement for research synthesis projects to conduct analyses to detect the presence of systematic outcome reporting biases. Some analyses go further by trying correct for its distorting effects on average effect size estimates. Widely known analytic techniques for doing so include rank-correlation tests, the Trim-and-Fill technique proposed by [Duval and Tweedie](https://dx.doi.org/10.2307/2669529), and [Egger regression](https://dx.doi.org/10.1136/bmj.315.7109.629) (in its [many variants](https://dx.doi.org/10.1186/1471-2288-9-2)). Another class of methods involves selection models (or weight function models), as proposed by [Hedges and Vevea](https://dx.doi.org/10.1007/BF02294384), [Vevea and Woods](https://dx.doi.org/10.1037/1082-989X.10.4.428), and others. As far as I can tell, these are well known among methodologists but very seldom applied due to their complexity and lack of ready-to-use software (though [an R package](https://CRAN.R-project.org/package=weightr) has recently become available).  More recent proposals include the PET-PEESE technique introduced by [Stanley and Doucouliagos](https://dx.doi.org/10.1002/jrsm.1095); Simonsohn, Nelson, and Simmon's [p-curve technique](https://dx.doi.org/10.1177/1745691614553988); Van Assen, Van Aert, and Wichert's [p-uniform](http://dx.doi.org/10.1037/met0000025), and others. The list of techniques grows by the day.

Among these methods, Egger regression, PET, and PEESE are superficially quite appealing due to their simplicity. These methods each involve estimating a fairly simple meta-regression model, using as the covariate the sampling variance of the effect size or some transformation thereof. PET uses the standard error of the effect size as the regressor; PEESE uses the sampling variance (i.e., the squared standard error); PET-PEESE involves first testing whether the PET estimate is statistically significant, using PEESE if it is or PET otherwise. The intercept from one of these regressions is the average effect size estimate from a study with zero sampling variance; the estimated intercept is used as a "bias-corrected" estimator of the population average effect. These methods are also appealing due to their extensibility. Because they are just meta-regressions, it is comparatively easy to extend them to meta-regression models that control for further covariates, to robust variance estimation to account for dependencies among effect size estimates, etc.

In [a recent blog post](http://datacolada.org/59), Uri Simonsohn reports some simulation evidence indicating that the PET-PEESE estimator can have large biases under certain conditions, _even in the absence of publication bias_. The simulations are based on standardized mean differences from two-group experiments and involve simulating collections of studies that include many with small sample sizes, as might be found in certain areas of psychology. On the basis of these performance assessments, he argues that this purported cure is actually worse than the disease---that PET-PEESE should _not_ be used in meta-analyses of psychological research because it performs to poorly to be trusted. In [a response to Uri's post](http://datacolada.org/wp-content/uploads/2017/04/Response-by-Joe-Hilgard-to-Colada-59.pdf), [Joe Hilgard](http://crystalprisonzone.blogspot.com/) suggests that some simple modifications to the method can improve its performance. Specifically, Joe suggests using a function of sample size as the covariate (in place of the standard error or sampling variance of $d$), and also using PET or PEESE as stand-alone estimators, rather than using them contingent on a significance test. 

In this post, I follow up Joe's suggestions while replicating and expanding upon Uri's simulations, to try and provide a fuller picture of the relative performance of these estimators. In brief, the simulations show that:

* In the absence of publication bias, the sample-size variants of PET and PEESE are __*exactly unbiased*__.
* However, these adjusted estimators still have a cost, being less precise than conventional fixed-effect or random-effects estimators. 
* In the presence of varying degrees of publication bias, the sample-size variant of PEESE does better than PET-PEESE as it is usually applied. But it's still not great. 

# Why use sample size?

To see why it makes sense to use a function of sample size as the covariate for PET-PEESE analyses, rather than using the standard error of the effect size estimate, let's look at the formulas. Say that we have a standardized mean difference estimate from a two-group design (without covariates) with sample sizes $n_0$ and $n_1$:
$$
d = \frac{\bar{y}_1 - \bar{y}_0}{s_p},
$$
where $\bar{y}_0$ and $\bar{y}_1$ are the sample means within each group and $s_p^2$ is the pooled sample variance. Following convention, we'll assume that the outcomes are normally distributed within each group, and the groups have common variance. The exact sampling variance of $d$ is a rather complicated formula, but which can be approximated reasonably well as
$$
\text{Var}(d) \approx \frac{n_0 + n_1}{n_0 n_1} + \frac{\delta^2}{2(n_0 + n_1)},
$$
where $\delta$ is the _true_ standardized mean difference parameter. This formula is a delta-method approximation. The first term captures the variance of the numerator of $d$, so it gets at how precisely the _unstandardized_ difference in means is estimated. The second term captures the variance of the denominator of $d$, so it gets at how precisely the _scale_ of the outcome is estimated. The second term also involves the unknown parameter $\delta$, which must be estimated in practice. The conventional formula for the estimated sampling variance of $d$ substitutes $d$ in place of $\delta$:
$$
V = \frac{n_0 + n_1}{n_0 n_1} + \frac{d^2}{2(n_0 + n_1)}.
$$

In PET-PEESE analysis, $V$ or its square root is used as a covariate in a regression of the effect sizes, as a means of adjusting for publication bias. There are two odd things about this. First, publication bias is about the statistical significance of the group differences, but statistical significance __*does not depend on the scale of the outcome*__. The test of the null hypothesis of no differences between groups is __*not*__ based on $d / \sqrt{V}$. Instead, it is a function of the $t$ statistic:
$$
t = d / \sqrt{\frac{n_0 + n_1}{n_0 n_1}}.
$$
Consequently, it makes sense to use __*only the first term of $V$*__ as a covariate for purposes of detecting publication biases. 

The second odd thing is that $V$ is generally going to be correlated with $d$ because we have to use $d$ to calculate $V$, as [Joe noted in his response to Uri](http://datacolada.org/wp-content/uploads/2017/04/Response-by-Joe-Hilgard-to-Colada-59.pdf). This means that there will be a non-zero correlation between $d$ and $V$ (or between $d$ and $\sqrt{V}$) except in some very specific cases, even in the absence of any publication bias. Pretty funky. 

This problem with regression tests for publication bias has been recognized for a while in the literature (e.g., Peters et al., 2006; Moreno, 2009), but most of the work here has focused on other effect size measures, like odds ratios, that are relevant in clinical medicine. 

# Simulation model

# Results

```{r setup, include = FALSE}
library(knitr)
opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
```


```{r cleaning}
library(stringr)
library(tidyr)
library(dplyr)
library(ggplot2)
library(trelliscopejs)

load("../files/PET-PEESE Simulation Results.Rdata")

results <- 
  results %>%
  select(-reps, -seed) %>%
  mutate(
    study_dist = ifelse(na == nb, "Uniform distribution of studies", 
                        ifelse(na > nb, "More small studies", "More large studies")),
    study_dist = factor(study_dist, 
                        levels = c("More small studies","Uniform distribution of studies","More large studies"))
  ) %>%
  unnest() %>%
  gather("estimator","val", ends_with("_M"), ends_with("_V")) %>%
  separate(estimator, c("estimator","stat"), sep = -2) %>%
  spread(stat, val) %>%
  mutate(
    estimator = str_replace(str_sub(estimator, 1, -2), "_","-"),
    estimator = factor(estimator, levels = c("FE-meta","RE-meta","Top-10","PET","PEESE","PET-PEESE","SPET","SPEESE","SPET-SPEESE")),
    RMSE = sqrt((M - mean_effect)^2 + V)
  )


bias_plot <- function(dat) {
  ggplot(dat, aes(mean_effect, M, color = estimator, shape = estimator)) + 
    geom_point() + geom_line() + 
    geom_abline(slope = 1, intercept = 0) + 
    facet_grid(p_RR ~ sd_effect) + 
    theme_light()
}

selected_estimators <- c("FE-meta","Top-10","PET-PEESE","SPEESE","SPET","SPET-SPEESE")

results %>%
  filter(study_dist == "Uniform distribution of studies", 
         studies == 100, n_max == 50,
         estimator %in% selected_estimators) %>%
  bias_plot()

```

```{r, eval=FALSE}

plot_dat <- 
  results %>%
  select(-n_min, -na, -nb, -V) %>%
  rename(maximum_n = n_max) %>%
  filter(estimator %in% selected_estimators) %>%
  group_by(study_dist, studies, maximum_n) %>%
  summarise(panel = panel(bias_plot(.)))

plot_dat %>%
  trelliscope(name = "Estimator expectation", nrow = 1, ncol = 1, self_contained = TRUE)
```

