---
title: The clubSandwich package for meta-analysis with RVE
author: 'James'
date: '2015-07-10'
slug: clubSandwich-for-RVE-meta-analysis
categories: []
tags:
  - meta-analysis
  - sandwiches
  - R
header:
  caption: ''
  image: ''
---

I've recently been working on small-sample correction methods for hypothesis tests in linear regression models with cluster-robust variance estimation. My colleague (and grad-schoolmate) Beth Tipton has developed small-sample adjustments for t-tests (of single regression coefficients) in the context of meta-regression models with robust variance estimation, and together we have developed methods for multiple-contrast hypothesis tests. We have an R package (called `clubSandwich`) that implements all this stuff, not only for meta-regression models but also for other models and contexts where cluster-robust variance estimation is often used.

The alpha-version of the package is currently [available on Github](https://github.com/jepusto/clubSandwich). See the Github README for instructions on how to install it in R. Below I demonstrate how to use the package to get robust variance estimates, t-tests, and F-tests, all with small-sample corrections. The example uses a dataset of effect sizes from a Campbell Collaboration [systematic review of dropout prevention programs](http://www.campbellcollaboration.org/lib/project/158/), conducted by Sandra Jo Wilson and her colleagues.

The original analysis included a meta-regression with covariates that capture methodological, participant, and program characteristics. I'll use a regression specification that is similar to Model III from Wilson et al. (2011), but treat the `evaluator_independence` and `implementation_quality` variables as categorical rather than interval-level; the original analysis clustered at the level of the sample (some studies reported results from multiple samples), whereas I will cluster at the study level. 
I fit the model two ways, first using the `robumeta` package and then using `metafor`.

#### robumeta model

```{r}
options(width=150)
library(robumeta)
library(clubSandwich)
data(dropoutPrevention)

m3_robu <- robu(LOR1 ~ study_design + attrition + group_equivalence + adjusted
                + outcome + evaluator_independence
                + male_pct + white_pct + average_age
                + implementation_quality + program_site + duration + service_hrs, 
                data = dropoutPrevention, studynum = studyID, var.eff.size = varLOR, 
                modelweights = "HIER")
print(m3_robu)
```

Note that `robumeta` produces small-sample corrected standard errors and t-tests, and so there is no need to repeat those calculations with `clubSandwich`. The `evaluator_independence` variable has four levels, and it might be of interest to test whether the average program effects differ by the degree of evaluator independence. The null hypothesis in this case is that the 10th, 11th, and 12th regression coefficients are all equal to zero. A small-sample adjusted F-test for this hypothesis can be obtained as follows. 
(The `vcov = "CR2"` option means that the standard errors will be corrected using the bias-reduced linearization method proposed by McCaffrey, Bell, and Botts, 2001.)

```{r}
Wald_test(m3_robu, constraints = 10:12, vcov = "CR2")
```

By default, the `Wald_test` function provides an F-type test with degrees of freedom estimated using the approximate Hotelling's $T^2_Z$ method. The test has less than 17 degrees of freedom, even though there are 152 independent studies in the data, and has a p-value of .07, so not-quite-significant at conventional levels. The low degrees of freedom are a consequence of the fact that one of the levels of `evaluator independence` has only a few effect sizes in it:

```{r}
table(dropoutPrevention$evaluator_independence)
```

#### metafor model

Our package also works with models fit using the `metafor` package. Here I re-fit the same regression specification, but use REML to estimate the variance components (`robumeta` uses a method-of-moments estimator) and use a somewhat different weighting scheme than that used in `robumeta`. 
```{r, message = FALSE}
library(metafor)
m3_metafor <- rma.mv(LOR1 ~ study_design + attrition + group_equivalence + adjusted
                      + outcome + evaluator_independence
                      + male_pct + white_pct + average_age
                      + implementation_quality + program_site + duration + service_hrs, 
                      V = varLOR, random = list(~ 1 | studyID, ~ 1 | studySample),
                     data = dropoutPrevention)
summary(m3_metafor)
```

`metafor` produces model-based standard errors, t-tests, and confidence intervals. The `coef_test` function from `clubSandwich` will calculate robust standard errors and robust t-tests for each of the coefficients:
```{r}
coef_test(m3_metafor, vcov = "CR2")
```
Note that `coef_test` assumed that it should cluster based on `studyID`, which is the outer-most random effect in the metafor model. This can also be specified explicitly by including the option `cluster = dropoutPrevention$studyID` in the call. 

The F-test for degree of evaluator independence uses the same syntax as before:
```{r}
Wald_test(m3_metafor, constraints = 10:12, vcov = "CR2")
```
Despite some differences in weighting schemes, the p-value is very close to the result obtained using `robumeta`.
