---
title: A handmade clubSandwich for multi-site trials
authors:
- admin
date: '2019-03-09'
draft: true
slug: clustered-and-interacted
categories: []
tags:
  - sandwiches
  - robust variance estimation
  - econometrics
  - weighting
  - instrumental variables
header:
  caption: ''
  image: ''
---


```{r, message = FALSE}
library(tidyverse)
data(STAR, package = "AER")

STAR_urban <-
  STAR %>%
  filter(
    # limit to urban/inner city schools
    schoolk %in% c("urban","inner-city"),
    # limit to complete outcome data
    !is.na(readk), !is.na(mathk)
  ) %>%
  droplevels() %>%
  # collapse control conditions
  mutate(
    stark = fct_collapse(stark, regular = c("regular","regular+aide"))
  ) %>%
  # calculate inverse-propensity weight
  group_by(schoolidk) %>%
  mutate(
    n = n(),
    nT = sum(stark=="small"),
    wt = ifelse(stark=="small", n / nT, n / (n - nT))
  ) %>%
  select(schoolidk, stark, readk, mathk, wt)

STAR_summary <- 
  STAR_urban %>%
  count(schoolidk)

STAR_urban %>%
  group_by(schoolidk, stark) %>%
  summarise(
    n = n(),
    wt = sum(wt)
  ) %>%
  mutate(n = sum(n)) %>%
  spread(stark, wt)

```
After these exclusions, the data include a total of `r sum(STAR_summary$n)` students from `r nrow(STAR_summary)` schools, ranging in size from `r min(STAR_summary$n)` to `r max(STAR_summary$n)` students.

For starters, let's get the average impacts using a seeming unrelated regression specification, with both conventional and clubSandwich standard errors. 
```{r}
library(clubSandwich)
```

```{r}
STAR_wt <- lm(cbind(readk, mathk) ~ 0 + schoolidk + stark, weights = wt, data = STAR_urban)

# conventional SEs
CR0 <- 
  coef_test(STAR_wt, vcov = "CR0", 
            cluster = STAR_urban$schoolidk, 
            test = "z",
            coefs = c("readk:starksmall","mathk:starksmall"))
CR0

# clubSandwich SEs
CR2 <- 
  coef_test(STAR_wt, vcov = "CR2", 
            cluster = STAR_urban$schoolidk, 
            coefs = c("readk:starksmall","mathk:starksmall"))

CR2
```

Now I'll do it "by hand"---or rather, with a bit of `dplyr`:

```{r}

# summary statistics by site

school_summaries <- 
  STAR_urban %>%
  group_by(schoolidk, stark) %>%
  summarise(
    # means by arm and site
    readk = mean(readk),
    mathk = mean(mathk),
    n_arm = n()
  ) %>%
  summarise(
    # impact estimates by site
    readk = diff(readk),
    mathk = diff(mathk),
    n = sum(n_arm),
    p = n_arm[stark=="small"] / n
  ) %>%
  mutate(
    w = n
  )

# overall impacts

school_summaries %>%
  gather("subject","impact_j", readk, mathk) %>%
  group_by(subject) %>%
  summarise(
    impact = weighted.mean(impact_j, w = w),
    SE_CR0 = sqrt(sum(w^2 * (impact_j - impact)^2) / sum(w)^2),
    SE_CR2 = sqrt(sum(w^2 * (impact_j - impact)^2 / (1 - w / sum(w))) / sum(w)^2),
    df_CR2 = 1 / (sum(w^2 / (sum(w) - w)^2) - 
                    2 * sum(w^3 / (sum(w) - w)^2) / sum(w) + 
                    sum(w^2 / (sum(w) - w))^2 / sum(w)^2)
  ) %>%
  knitr::kable(digits = 2)
```

The CR0 and CR2 standard errors match the results from `coef_test`, as do the Satterthwaite degrees of freedom. Note that the degrees of freedom are equal to `r round(CR2$df[1], 1)` in this case, a bit less than $J - 1 = 22$ due to variation in the weight assigned to each school. 

