---
title: 'Converting from odds ratios to standardized mean differences: What to do with logistic regression coefficients?'
authors:
- admin
date: '2019-05-26'
slug: Converting-odds-ratios-to-standardized-mean-differences
categories: []
draft: true
tags:
  - meta-analysis
  - odds ratio
  - effect size
header:
  caption: ''
  image: ''
  preview: yes
---

One of the central problems in research synthesis is that studies use a variety of different types of outcome measures to assess a construct. This is the main reason that meta-analysis often uses standardized, scale-free effect sizes (such as standardized mean differences), so that findings from studies that use different measures can be combined and contrasted on a common metric. In syntheses of education research (as well as other fields), a further issue that sometimes arises is that some included studies might report effects on a dichotomous outcome, while others report effects (of the same intervention, say) but using a continuous outcome measure. 

```{r}
logit <- function(x) log(x) - log(1 - x)

simulate_OR_to_SMD <- function(p0, SMD, r, n0, n1) {
  
  # simulate data
  trt <- c(rep(0, n0), rep(1, n1))
  Y <- rlogis(n0 + n1, location = logit(p0) + trt * SMD * pi / sqrt(3))
  X <- r * (Y - trt * SMD) * sqrt(3) / pi + rnorm(n0 + n1, sd = sqrt(1 - r^2))
  B <- Y > 0

  # calculate LORs
  logit_fit <- glm(B ~ trt + X, family = "binomial")
  LOR_marginal <- as.numeric(diff(logit(tapply(B, trt, mean))))
  LOR_logit <- coef(logit_fit)[["trt"]]
  LORs <- c(LOR_marginal, LOR_logit)
  
  # convert to SMDs
  SMDs <- LORs * sqrt(3) / pi
  
  data.frame(type = c("marginal","conditional"), LOR_est = LORs, SMD_est = SMDs)
} 

simulate_OR_to_SMD(p0 = 0.6, SMD = 0.4, r = 0.7, n0 = 10000, n1 = 10000)

```

```{r}
library(tidyverse)

params <- 
  list(
    p0 = seq(0.2, 0.8, 0.2),
    SMD = seq(0.2, 0.8, 0.2),
    r = seq(0, 0.9, 0.1)
  ) %>%
  cross_df()

SMDs <- 
  params %>%
  mutate(res = pmap(., simulate_OR_to_SMD, n0 = 50000, n1 = 50000)) %>%
  unnest() %>%
  mutate(RB = SMD_est / SMD)

ggplot(SMDs, aes(r, RB, color = type)) + 
  geom_point() + geom_line() + 
  facet_grid(SMD ~ p0, labeller = "label_both") + 
  theme_light()
```

