---
title: Getting started with scdhlm
authors:
- admin
date: '2014-10-19'
slug: getting-started-with-scdhlm
categories: []
tags:
  - single-case design
  - design-comparable SMD
  - Rstats
header:
  caption: ''
  image: ''
---

**UPDATED 10/2/2016 after posting the package to CRAN**

Here are step-by-step instructions on how to download and install the scdhlm package for R. You'll need to have a [copy of R installed](http://cran.us.r-project.org/). There are two ways to do the installation: through the Comprehensive R Archive Network (CRAN) or from the source code on Github. I describe each approach in turn.

### Option 1: Via CRAN

Go via CRAN to install the most recent stable version of the package. Type the following commands at the R prompt:

```{r eval = FALSE}
install.packages("scdhlm")
library(scdhlm)
```

### Option 2: Via Github

Go via Github to get the latest development version of the package. For this option, you will first need to install the devtools package:

```{r, eval=FALSE}
install.packages("devtools")
```

Once you have successfully installed this package, type the following:

```{r, eval=FALSE}
library(devtools)
install_github("jepusto/scdhlm")
library(scdhlm)
```

### Further instructions

You'll only need to do the installation once. Once you've got the package installed, type the following in order to access the package within an R session: `library(scdhlm)`. 

To open the package documentation, type `package?scdhlm`. To access the documentation for an individual function in this package, just type `?` followed by the name of the function. For instance, one of the main functions in the package is called `g_REML`; to access its documentation, type `?g_REML`.

### web-interface for calculating effect sizes

The package includes an interactive app (written with `shiny`) for calculating design-comparable standardized mean differences. To run this app on your computer, you will first need to [install RStudio](https://www.rstudio.com/products/rstudio/download/) (if you don't already have it). Then ensure that you have the `shiny`, `markdown`, and `ggplot2` packages installed by running the following:  
```{r, eval=FALSE}
install.packages("shiny")
install.packages("markdown")
install.packages("ggplot2")
```

Finally, open the app by typing the following at the prompt within RStudio:
```{r, eval=FALSE}
library(scdhlm)
shine_scd()
```

The app should now open in your web browser. 
