---
title: Another meta-sandwich
author: 'James'
date: '2014-04-23'
slug: Robust-meta-analysis-2
categories: []
tags:
  - meta-analysis
  - sandwiches
  - R
header:
  caption: ''
  image: ''
---

In [a previous post](/Robust-meta-analysis-1/), I provided some code to do robust variance estimation with `metafor` and `sandwich`. 
Here's another example, replicating some more of the calculations from [Tanner-Smith & Tipton (2013)](http://doi.org/10.1002/jrsm.1091). 
([See here](https://gist.github.com/jepusto/11147304) for the complete code.)

As a starting point, here are the results produced by the `robumeta` package:
```{r}
library(grid)
library(robumeta)

data(corrdat)
rho <- 0.8

HTJ <- robu(effectsize ~ males + college + binge,
            data = corrdat, 
            modelweights = "CORR", rho = rho,
            studynum = studyid,
            var.eff.size = var, small = FALSE)
HTJ
```

To exactly re-produce the results with `metafor`, I'll need to use the weights proposed by HTJ. In their approach to the correlated effects case, effect size $i$ from study $j$ receives weight equal to $\left[\left(v_{\cdot j} + \hat\tau^2\right)(1 + (k_j - 1) \rho)\right]^{-1}$, where $v_{\cdot j}$ is the average sampling variance of the effect sizes from study $j$, $\hat\tau^2$ is an estimate of the between-study variance, $k_j$ is the number of correlated effects in study $j$, and $\rho$ is a user-specified value of the intra-study correlation. However, it appears that `robumeta` actually uses a slightly different set weights, which are equivalent to taking $\rho = 1$. I calculate the latter weights, fit the model in `metafor`, and output the robust standard errors and $t$-tests:

```{r, message=FALSE}
devtools::source_gist(id = "11144005", filename = "metafor-sandwich.R")

corrdat <- within(corrdat, {
  var_mean <- tapply(var, studyid, mean)[studyid]
  k <- table(studyid)[studyid]
  var_HTJ <- as.numeric(k * (var_mean + as.numeric(HTJ$mod_info$tau.sq)))
})

meta1 <- rma.mv(effectsize ~ males + college + binge, 
                V = var_HTJ, 
                data = corrdat, method = "FE")
meta1$cluster <- corrdat$studyid
RobustResults(meta1)
```

One could specify a similar (though not exactly identical model) in `metafor` as follows. In the HTJ approach, $\rho$ represents the total correlation induced by both the within-study sampling error and intra-study correlation in true effects. In contrast, the `metafor` approach would take $\rho$ to be correlation due to within-study sampling error alone. I'll first need to create a block-diagonal covariance matrix given a user-specified value of $\rho$. 
```{r}
library(Matrix)
equicorr <- function(x, rho) {
  corr <- rho + (1 - rho) * diag(nrow = length(x))
  tcrossprod(x) * corr 
} 
covMat <- as.matrix(bdiag(with(corrdat, tapply(var_mean, studyid, equicorr, rho = 0.8, simplify = FALSE))))
```

Passing this block-diagonal covariance matrix to `rma.mv`, I now estimate the model 

$$T_{ij} = \mathbf{X}_{ij} \beta + \nu_i + e_{ij},$$

where $Var(\nu_i) = \sigma^2$, $Var(e_{ij}) = v_{ij}$, and $Cor(e_{ij}, e_{ik}) = \rho$. Note that $\sigma^2$ is now estimated via REML.

```{r}
meta2 <- rma.mv(yi = effectsize ~ males + college + binge, 
                V = covMat, random = ~ 1 | studyid, 
                data = corrdat,
                method = "REML")
c(sigma.sq = meta2$sigma2)
```

The between-study heterogeneity estimate is considerably larger than the moment estimate from `robumeta`. Together with the difference in weighting, this leads to some changes in the coefficient estimates and their estimated precision:

```{r}
RobustResults(meta2)
```

It is important to keep in mind that the estimate of between-study heterogeneity depends on the posited model for the covariance structure, including the assumed value of $\rho$. HTJ recommend conducting sensitivity analysis across a range of values for the within-study effect correlation. Re-calculating the value of $\sigma^2$ for $\rho$ between 0.0 and 0.9 yields the following:

```{r}
sigma2 <- function(rho) {
  covMat <- as.matrix(bdiag(with(corrdat, tapply(var_mean, studyid, equicorr, rho = rho, simplify = FALSE))))
  rma.mv(yi = effectsize ~ males + college + binge, 
                  V = covMat, random = ~ 1 | studyid, 
                  data = corrdat,
                  method = "REML")$sigma2
}
rho_sens <- seq(0,0.9,0.1)
sigma2_sens <- sapply(rho_sens, sigma2)
cbind(rho = rho_sens, sigma2 = round(sigma2_sens, 4))
```

The between-study heterogeneity is quite insensitive to the assumed value of $\rho$. 

The difference between the results based on `metafor` versus on `robumeta` appears to be due to the subtle difference in the weighting approach: `metafor` uses block-diagonal weights that contain off-diagonal terms for effects drawn from a common study, whereas `robumeta` uses entirely diagonal weights.
